Abstract Data Type Inference

### 2015-03-25 formula.tex
some final typos

\section{Computing Symbolic Representations}
\label{sec:formula}

TODO MOTIVATE THIS SECTION

Let $h = \tup{O,<,c,f,m,r}$ be a history with operations $O = \set{ o_1, \ldots,
o_n }$. Without loss of generality, we assume the match targets $\set{o_1,
\ldots, o_k}$ of $h$ are indexed consecutively from $1$ to $k$, for some $k \le
n$. For each $o \in O$, we define the macro $\textsc{Embed}_o(x,Y,z)$:
\begin{align*}
  & \big(c(o) \implies {\sf c}(x)\big) \land {\sf f}(x) = f(o) \land {\sf r}(x) = r(o) \\
  & \land \big(o \in \dom(m)
    \implies \lnot{\sf um}(x) \land {\sf m}(x) = z\big) \\
  & \land \big(o \not\in \dom(m) \implies {\sf um}(x)\big)
    \land \bigwedge_{y \in Y} x < y
\end{align*}
capturing the correspondence between the operation $o$ and the logical variable
$x$ representing $o$. The variables $Y$ and $z$ represent the operations
ordered after $x$, and the operation which $x$ matches. The macro
$\textsc{Identical}(x,y)$:
\begin{align*}
  & \big({\sf c}(x) \Leftrightarrow {\sf c}(y)\big)
    \land {\sf f}(x) = {\sf f}(y) 
    \land \big({\sf r}(x) \Leftrightarrow {\sf r}(y)\big) \\
  & \land \big({\sf um}(x) \Leftrightarrow {\sf um}(y)\big)
    \land {\sf m}(x) = {\sf m}(y)
\end{align*}
captures whether the operations bound to $x$ and $y$ are identical.
To express the constraints among the matches of embedded operations, we define
the macro $\textsc{Match}(Y,z)$:
\begin{align*}
  \forall x.
  {\sf m}(x) = z \implies
  {\sf r}(x) \lor \bigvee_{y \in Y} \textsc{Identical}(x,y)
\end{align*}
which requires any operation which matches $z$ to be either read-only, or
identical to some operation in $Y$, which represents the operations of $h$
which match $z$. Finally, we express the embedding of $h$ with the macro
$\textsc{Embed}_h$:
\begin{align*}
  \exists x_1, \ldots, x_n.
  \bigwedge_{i=1}^{n} \textsc{Embed}_{o_i}(x_i,Y_i,z_i)
  \land \bigwedge_{i=1}^{k} \textsc{Match}(W_i, x_i)
\end{align*}
where $Y_i = \set{ x_j : o_i < o_j }$ are the variables corresponding to
operations ordered after $o_i$, and $z_i$ is the variable corresponding to
$m(o_i)$, if defined, and $W_i = \set{ x_j : m(o_j) = o_i }$ are the variables
corresponding to operations matching $o_i$.

@@

  Consider again the histories of Example~\ref{ex:patterns} which generate the
  complement of the atomic register ADT. The $\textsc{Embed}$ formula for the
  first history,
\begin{verbatim}
  [1:X] read => 1 (RO)  #
\end{verbatim}
  after simplifications, like replacing $\mathrm{true} \implies p$ with $p$, is
  \begin{align*}
    \exists x_1.\ {\sf c}(x_1) \land {\sf f}(x_1) = \mathrm{read} \land {\sf um}(x_1).
  \end{align*}
  The $\textsc{Embed}$ formula for second history,
\begin{verbatim}
  [1:1] write(1)            #
  [2:2] read => empty (RO)    #    
\end{verbatim}
  is similarly given by
  \begin{align*}
    & \exists x_1, x_2.\ x_1 < x_2 \\
    & \quad \land {\sf c}(x_1) \land {\sf f}(x_1) = \mathrm{write}
      \land \lnot{\sf um}(x_1) \land {\sf m}(x_1) = x_1 \\
    & \quad \land {\sf c}(x_2) \land {\sf f}(x_2) = \mathrm{read}
      \land \lnot{\sf um}(x_2) \land {\sf m}(x_2) = x_2 \\
    & \quad \land \big(\forall x.\ {\sf m}(x) = x_1 \implies {\sf r}(x) \lor \textsc{Identical}(x,x_1)\big) \\
    & \quad \land \big(\forall x.\ {\sf m}(x) = x_2 \implies {\sf r}(x) \lor \textsc{Identical}(x,x_2)\big).
  \end{align*}
  The formulas for the other histories are similarly obtained.

@@

  The formula $\bigwedge_{h \in H} \neg \textsc{Embed}_h$ represents $A$,
  for any set $H$ which generates the compliment of $A$.

>>>>

%!TEX root = draft.tex
\section{Computing Symbolic Representations}
\label{sec:formula}

While Section~\ref{sec:patterns} demonstrates that (the kernel complements) of
naturally-occurring ADTs can be represented finitely, in this section we
demonstrate that such finite representations have logical interpretations,
allowing us to derive the formulas representing ADTs. In what follows, we
describe how to obtain a formula which is satisfied by the histories which
embed a given history, via $\preceq$. Then, using the finite set of histories
which represent the complement of a given ADT~\footnote{The complement of the kernel of an ADT $A$ is simply called the complement of $A$.}, we represent the ADT itself as
the conjunction of negations of these embedding formulas. The resulting formula
is satisfied only on the histories which do not embed the generators of a given
ADT’s complement.

Let $h = \tup{O,<,c,f,m,r}$ be a history with operations $O = \set{ o_1, \ldots,
o_n }$. Without loss of generality, we assume the match targets $\set{o_1,
\ldots, o_k}$ of $h$ are indexed consecutively from $1$ to $k$, for some $k \le
n$. For each $o \in O$, we define the macro $\textsc{Embed}_o(x,Y,z)$:
\begin{align*}
  & \big(c(o) \implies {\sf c}(x)\big) \land {\sf f}(x) = f(o) \land {\sf r}(x) \Leftrightarrow r(o) \\
  & \land \big(o \in \dom(m)
    \implies \lnot{\sf um}(x) \land {\sf m}(x) = z\big) \\
  & \land \big(o \not\in \dom(m) \implies {\sf um}(x)\big)
    \land \bigwedge_{y \in Y} x < y
\end{align*}
capturing the correspondence between the operation $o$ and the logical variable
$x$ representing $o$. The variables $Y$ and $z$ represent the operations
ordered after $x$, and the operation which $x$ matches. The macro
$\textsc{Identical}(x,y)$:
\begin{align*}
  & \big({\sf c}(x) \Leftrightarrow {\sf c}(y)\big)
    \land {\sf f}(x) = {\sf f}(y) 
    \land \big({\sf r}(x) \Leftrightarrow {\sf r}(y)\big) \\
  & \land \big({\sf um}(x) \Leftrightarrow {\sf um}(y)\big)
    \land {\sf m}(x) = {\sf m}(y)
\end{align*}
captures whether the operations bound to $x$ and $y$ are identical.
To express the constraints among the matches of embedded operations, we define
the macro $\textsc{Match}(Y,z)$:
\begin{align*}
  \forall x.
  {\sf m}(x) = z \implies
  {\sf r}(x) \lor \bigvee_{y \in Y} \textsc{Identical}(x,y)
\end{align*}
which requires any operation which matches $z$ to be either read-only, or
identical to some operation in $Y$, which represents the operations of $h$
which match $z$. Finally, we express the embedding of $h$ with the macro
$\textsc{Embed}_h$:
\begin{align*}
  \exists x_1, \ldots, x_n.
  \bigwedge_{i=1}^{n} \textsc{Embed}_{o_i}(x_i,Y_i,z_i)
  \land \bigwedge_{i=1}^{k} \textsc{Match}(W_i, x_i)
\end{align*}
where $Y_i = \set{ x_j : o_i < o_j }$ are the variables corresponding to
operations ordered after $o_i$, and $z_i$ is the variable corresponding to
$m(o_i)$, if defined, and $W_i = \set{ x_j : m(o_j) = o_i }$ are the variables
corresponding to operations matching $o_i$.

@@

  Consider again the histories of Example~\ref{ex:patterns} which generate the
  complement of the atomic register ADT. The $\textsc{Embed}$ formula for the
  first history,
\begin{verbatim}
  [1:X] read => 1 (RO)  #
\end{verbatim}
  after simplifications, like replacing $\mathrm{true} \implies p$ with $p$, is
  \begin{align*}
    \exists x_1.\ {\sf c}(x_1) \land {\sf f}(x_1) = \mathrm{read} \land {\sf um}(x_1) \land {\sf r}(x_1).
  \end{align*}
  The $\textsc{Embed}$ formula for second history,
\begin{verbatim}
  [1:1] write(1)            #
  [2:2] read => - (RO)    #    
\end{verbatim}
  is similarly given by
  \begin{align*}
    & \exists x_1, x_2.\ x_1 < x_2 \\
    & \quad \land {\sf c}(x_1) \land {\sf f}(x_1) = \mathrm{write}
      \land \lnot{\sf um}(x_1) \land {\sf m}(x_1) = x_1 \\
    & \quad \land {\sf c}(x_2) \land {\sf f}(x_2) = \mathrm{read}
      \land \lnot{\sf um}(x_2) \land {\sf m}(x_2) = x_2\land {\sf r}(x_2) \\
    & \quad \land \big(\forall x.\ {\sf m}(x) = x_1 \implies {\sf r}(x) \lor \textsc{Identical}(x,x_1)\big) \\
    & \quad \land \big(\forall x.\ {\sf m}(x) = x_2 \implies {\sf r}(x) \lor \textsc{Identical}(x,x_2)\big).
  \end{align*}
  The formulas for the other histories are similarly obtained.

@@

  The formula $\bigwedge_{h \in H} \neg \textsc{Embed}_h$ represents $A$,
  for any set $H$ which generates the complement of $A$.

### 2015-03-25 histories.tex
some final typos

\end{itemize}
Non-completed operations are \emph{pending}, and are maximal in the
happens-before order.

@@

\end{itemize} 
The \emph{frequency} of an operation $o$ is the number of operations identical
to $o$, denoted $\freq(o)$. We say that $o$ has \emph{duplicates} when
$\freq(o) > 1$. An operation $o \in \img(m)$ in the image of $m$ is a
\emph{match target}, and the inverse set $m^{-1}(o)$ of a target is a
\emph{match}. We assume $o \in m^{-1}(o)$. A operation $o$ is \emph{unmatched}
when $o$ is completed and $m(o)$ is undefined.

@@

    \item read $\Rightarrow v$ operations are read-only, and match themselves
    when they return $v = \mathord\perp$, or a write$(v)$ operation, if one
    exists, and otherwise have no match,

@@

  \item removing an order constraint (o),

>>>>

\end{itemize}
The relation $<$ is an interval order because histories abstract executions where
the set of call and return actions is totally ordered~\cite{conf/popl/BouajjaniEEH15}.
Non-completed operations are \emph{pending}, and are maximal in the
happens-before order.

@@

\end{itemize} 
The \emph{frequency} of an operation $o$ is the number of operations identical
to $o$, denoted $\freq(o)$. We say that $o$ has \emph{duplicates} when
$\freq(o) > 1$. An operation $o \in \img(m)$ in the image of $m$ is a
\emph{match target}, and the inverse set $m^{-1}(o)$ of a target is a
\emph{match}. We assume $o \in m^{-1}(o)$ for every $o$. A operation $o$ is \emph{unmatched}
when $o$ is completed and $m(o)$ is undefined.

@@

    \item read $\Rightarrow v$ operations are read-only, and match themselves
    when they return $v = -$, or a write$(v)$ operation, if one
    exists, and otherwise have no match,

@@

  \item unordering a pair of ordered operations (o),

### 2015-03-25 matching.tex
some final typos

\section{Computing Matching Schemes}
\label{sec:matching}

@@

We automate the computation of matching schemes by considering the function
${\sf g}$ underlying a simple matching scheme as an uninterpreted function in a
logical satisfiability problem which uniquely determines the match of each
operation, and constrain ${\sf g}$ to normalize the executions of an
implementation. To this end, we fix an implementation $\mathcal{I}$ and a
language $L = \tup{E,P,Q}$ to which $\mathcal{I}$ adheres, then consider any
enumeration $F$ of execution pairs $\tup{e,e'} \in E^2$ such that
\begin{itemize}

>>>>

%!TEX root = draft.tex
\section{Computing Matching Schemes}
\label{sec:matching}

@@

We automate the computation of matching schemes by constructing a logical
formula that characterizes the boolean functions $G$ underlying a simple 
matching scheme. These boolean functions are defined as the interpretation of 
a function symbol ${\sf g}$. 
%the function
%symbol ${\sf g}$ underlying a simple matching scheme as an uninterpreted function in a
%logical satisfiability problem which 
The formula expresses the fact that the interpretations of ${\sf g}$ uniquely determine the match of each
operation, and that they normalize the executions of an
implementation. To this end, we fix an implementation $\mathcal{I}$ and a
language $L = \tup{E,P,Q}$ to which $\mathcal{I}$ adheres, then consider any
enumeration $F$ of execution pairs $\tup{e,e'} \in E^2$ such that
\begin{itemize}

### 2015-03-25 nature.tex
Updated intro.
some final typos

\section{Properties of Naturally-Occurring ADTs}
\label{sec:nature}

@@

  \item read $\Rightarrow v$ returns the last-stored value $v$, or the nil
  value $\perp$ if no value has yet been stored.

@@

  \item read $\Rightarrow v$ operations are read-only, and match themselves if
  $v = \perp$, or the unique write$(v)$ operation, if one exists, and otherwise 
  have no match.

@@

It is easy to see that the histories of atomic registers normal. They are
closed under removal of read-only and duplicate operations, since the each
match can contain an arbitrary number of read operations. Since the histories
of atomic registers do not contain unmatched operations, they are also closed
under their removal. Finally, since the kernel contains an arbitrary number of
sequenced matches, atomic register histories are also closed under match
removal.

Our inference algorithm computes following four histories to generate the
complement of the atomic register ADT:
\begin{verbatim}
  [1:X] read => 1 (RO)  #        [1:1] write(1)        #
  ---                            [2:2] read => - (RO)    #
  [1:1] write(1)        #        ---
  [2:2] write(2)          #      [1:2] read => 1 (RO)  #
  [3:1] read => 1 (RO)      #    [2:2] write(1)          #
\end{verbatim}
As new generators are discovered for $n=1$ and $n=2$ matches only, the atomic
register ADT is predictable.

@@

  \item remove $\Rightarrow v$ returns the nil value $v = \mathord\perp$ if the
  collection is empty, and otherwise removes and returns the least- or
  most-recently added value $v$, respectively.

@@

  \item remove $\Rightarrow v$ operations are read-only, and match themselves
  if $v = \mathord\perp$ is the nil value. If $v \neq \mathord\perp$, remove
  $\Rightarrow v$ operations are not read-only, and match the unique add$(v)$
  operation, if one exists, and otherwise have no match.

@@

It is not hard to see that the histories of atomic queues and stacks are
normal. They are closed under removal of read-only operations: only empty
removes, i.e.,~remove $\Rightarrow \mathord\perp$, are read-only. They are
closed under the removal of duplicate operations: only non-empty removes can be
duplicates, and such duplicates are not admitted in the first place, since each
value is added only once. Similarly, unmatched operations, i.e.,~removes that
return values that have not been added, are not admitted in the first place.
Finally, since the removal of entire matches preserves the FIFO/LIFO behavior
of the entire collection, and the correctness of empty returns, histories are
also closed under match removal.

@@

\end{itemize}
Each operation returns the nil value $v = \mathord\perp$ when $u$ is not yet
present, and otherwise returns $v = u$. As its name implies, its kernel is
sequential.

An execution \emph{inserts unique values} if the argument value to each insert
operation which returns $\mathord\perp$ is unique. We give a faithful matching
scheme over executions which insert unique values as follows:
\begin{itemize}

  \item any operation returning $v = \mathord\perp$ matches itself, and

  \item any operation returning $v \neq \mathord\perp$ matches the unique
  insert$(u) \Rightarrow \mathord\perp$ operation, if one exists, and otherwise
  has no match.

@@

\end{itemize}
An operation returns the nil value $v = \mathord\perp$ when the lock is not
currently held, indicating success for lock, and failure for unlock. Otherwise,
the operations return the value $u = v$ passed as an argument to the last
successful lock$(v)$ operation.

@@

  \item any operation returning $v = \mathord\perp$ matches itself, and

  \item any operation returning $v \neq \mathord\perp$ matches the unique
  lock$(u) \Rightarrow \mathord\perp$ operation, if one exists, and otherwise
  has no match.

>>>>

%!TEX root = draft.tex
\section{Properties of Naturally-Occurring ADTs}
\label{sec:nature}

@@

  \item read $\Rightarrow v$ returns the last-stored value $v$, or the nil
  value $-$ if no value has yet been stored.

@@

  \item read $\Rightarrow v$ operations are read-only, and match themselves if
  $v = -$, or the unique write$(v)$ operation, if one exists, and otherwise 
  have no match.

@@

It is easy to see that the atomic register is normal. It is
closed under removal of read-only and duplicate operations, since  each
match can contain an arbitrary number of read operations. Since the histories
of atomic registers do not contain unmatched operations, they are also closed
under their removal. Finally, since the kernel contains an arbitrary number of
sequenced matches, atomic register histories are also closed under match
removal.

Our inference algorithm computes the following four histories to generate the
complement of the atomic register ADT:
\begin{verbatim}
  [1:X] read => 1 (RO)  #        [1:1] write(1)        #
  ---                            [2:2] read => - (RO)    #
  [1:1] write(1)        #        ---
  [2:2] write(2)          #      [1:2] read => 1 (RO)  #
  [3:1] read => 1 (RO)      #    [2:2] write(1)          #
\end{verbatim}
As new generators are discovered for $n=1$ and $n=2$ matches only, the atomic
register ADT is predictable.

@@

  \item remove $\Rightarrow v$ returns the nil value $v = -$ if the
  collection is empty, and otherwise removes and returns the least- or
  most-recently added value $v$, respectively.

@@

  \item remove $\Rightarrow v$ operations are read-only, and match themselves
  if $v = -$ is the nil value. If $v \neq -$, remove
  $\Rightarrow v$ operations are not read-only, and match the unique add$(v)$
  operation, if one exists, and otherwise have no match.

@@

It is not hard to see that the atomic queues and stacks are
normal. They are closed under removal of read-only operations: only empty
removes, i.e.,~remove $\Rightarrow -$, are read-only. They are
closed under the removal of duplicate operations: only non-empty removes can be
duplicates, and such duplicates are not admitted in the first place, since each
value is added only once. Similarly, unmatched operations, i.e.,~removes that
return values that have not been added, are not admitted in the first place.
Finally, since the removal of entire matches preserves the FIFO/LIFO behavior
of the entire collection, and the correctness of empty returns, histories are
also closed under match removal.

@@

\end{itemize}
Each operation returns the nil value $v = -$ when $u$ is not yet
present, and otherwise returns $v = u$. As its name implies, its kernel is
sequential.

An execution \emph{inserts unique values} if the argument value to each insert
operation which returns $-$ is unique. We give a faithful matching
scheme over executions which insert unique values as follows:
\begin{itemize}

  \item any operation returning $v = -$ matches itself, and

  \item any operation returning $v \neq -$ matches the unique
  insert$(v) \Rightarrow -$ operation, if one exists, and otherwise
  has no match.

@@

\end{itemize}
An operation returns the nil value $v = -$ when the lock is not
currently held, indicating success for lock, and failure for unlock. Otherwise,
the operations return the value $u = v$ passed as an argument to the last
successful lock$(u)$ operation.

@@

  \item any operation returning $v = -$ matches itself, and

  \item any operation returning $v \neq -$ matches the unique
  lock$(v) \Rightarrow -$ operation, if one exists, and otherwise
  has no match.
